\bf The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-05-10 paper/paper.Rmd
First pass at Conclusion -> Future Work

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. Furthermore, we also envision the package being used more generally for inventory plot-structure with many spatially separated plots, e.g. the US Forest Service Forest Inventory and Analysis plots [@smith_forest_2002]. In future versions of `forestecology` we also hope to include models that account for tree mortality in addition to tree growth. The package follows the `tidy` data design principles, leverages the `sf` package for spatial data, and S3 open-oriented model implementation structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

>>>>

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. The package follows the `tidy` data design principles, leverages the `sf` package for spatial data, and S3 open-oriented model implementation structure. We hope that the package will increase the use of neighborhood competition models to better understand what structures plant competition.

While the package is designed with ForestGEO plot data in mind, we envision that it can be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. Furthermore, we also envision the package being used more generally for inventory plot-structure with many spatially separated plots, e.g. the US Forest Service Forest Inventory and Analysis plots [@smith_forest_2002].

We hope to extend the `forestecology` package's functionality to account for a larger variety of models for tree growth and corresponding tests. Some immediate improvements that can made include

> To address the reviewer's critique that the model is not flexible enough and does not include applications of existing competition theory, we could add another option for run_shuffle. In this case the shuffle would not shuffle identity of competitors which share with the focal species. The idea being there is a special kind of competition with conspecifics (because they share your exact resource needs). The result is two lambda values, one for competition between conspecifics and one for competition between heterospecifics.

A recently emerging body of literature suggests that traits of species rather than species identity themselves better inform competition models [@kunstler_competitive_2012, @lasky_trait-mediated_2014]. To account for such clustering, instead of using Equation \ref{eq:model}, we present a model of which is a modification of @kunstler_plant_2016 in Equation \ref{eq:model2}. Let $Y_{ik}$ be the growth of the $k^{\text{th}}$ individual of species $i$ and let $j$ be the species of competitors in the neighborhood. How much growth changes per unit of $\text{BA}$ of all neighbors is quantified by two terms: an overall competition coefficient $C$ and an $i$-$j$ species pair specific coefficient $c_{ij}$. which depends on the traits of the $i$ and $j$ species. Say we have $l$ traits and $t_{li}$ is the value of trait $l$ for species $i$. $a_{\text{tol},l}$ quantifies how trait $l$ affects the way a species tolerates (or is sensitive to) competition. $a_{\text{eff},l}$ quantifies how trait $l$ is how much trait $l$ affects how strong a competitor is, how much it effects the reduction in growth of its neighbors. Finally $a_{\text{sim},l}$ is about whether competition is stronger (or weaker) if the two species are similar. This is based on an idea that two species that are similar use the same resources and should compete more.

\label{eq:model2}
\begin{eqnarray}
y_{ik} &=& \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{j} (C + c_{ij}) \cdot \text{BA}_{j} + \epsilon_{ij}\\
\text{where } c_{ij} &=& \sum_{l} a_{\text{tol},l}\cdot t_{li} + a_{\text{eff},l}\cdot t_{lj} + a_{\text{sim},l}|t_{li} - t_{lj}|\nonumber
\end{eqnarray}

Lastly, the `forestecology` current uses the `blockCV` package behind the scenes to create the spatial blocks acting as folds for our spatial cross-validation algorithm detailed in Sections \ref{spatial-information} and \ref{spatial-cross-validation} [@valavi_blockcv_2019]. This back-end functionality could be substituted with the `spatialsample` package for spatial resampling infrastructure; a `tidymodels` package under active development as of 2021 [@spatialsample_package, @tidymodels_package].

### 2021-05-10 paper/paper.tex
Switched image png -> pdf
First pass at Conclusion -> Future Work

\caption{Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot.}\label{fig:scbi-trees}
\end{figure}

@@

\caption{Add spatial information. A buffer region and spatial cross-validation blocks 1 througth 4. The location of each tree is marked with its fold number where the folds are delineated with solid lines. The color of each digit indicates whether the tree is part of the buffer region (thus will only be considered as a competitor tree) or is part of the interior of the study region (thus is a focal tree whose growth is of modeled interest).}\label{fig:scbi-spatial-information}
\end{figure}

@@

\caption{Identify all focal and corresponding competitor trees. The dashed circle extends 7.5m away from the focal tree 4 while all 20 competitor trees are within this circle.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}

@@

\caption{Fit model. Posterior distributions of all parameters. For compactness we include only three species.}\label{fig:scbi-posterior-viz}
\end{figure}

@@

This process is repeated for each of the four folds acting as the test
set, then the four RMSE's are averaged to provide a single estimate of
model error. This algorithm is implemented in \texttt{run\_cv()}, which
acts as a wrapper function to both \texttt{comp\_bayes\_lm()} that fits
the model and \texttt{predict()} that returns predicted values.

@@

\caption{Root mean squared error of models for standard, permuted, and spatial cross-validated error estimates. The dotted lines show non-permuted competitor identity, while the histogras so the RMSE for 49 permutations. The colors indicate whether cross validaton was used.}\label{fig:scbi-simulation}
\end{figure}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. While the package is designed
with ForestGEO plot data in mind, we envision that it can be modified to
work on any single large, mapped forest plot in which at least two
measurements of each individual have been taken. Furthermore, we also
envision the package being used more generally for inventory
plot-structure with many spatially separated plots, e.g.~the US Forest
Service Forest Inventory and Analysis plots \citep{smith_forest_2002}.
In future versions of \texttt{forestecology} we also hope to include
models that account for tree mortality in addition to tree growth. The
package follows the \texttt{tidy} data design principles, leverages the
\texttt{sf} package for spatial data, and S3 open-oriented model
implementation structure. We hope that the package will increase the use
of neighborhood competition models to better understand what structures
plant competition.

>>>>

\caption{Step 1 - Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot.}\label{fig:scbi-trees}
\end{figure}

@@

\caption{Step 2 - Add spatial information. A buffer region and spatial cross-validation blocks 1 through 4. The location of each tree is marked with its fold number where the folds are delineated with solid lines. The color of each digit indicates whether the tree is part of the buffer region (thus will only be considered as a competitor tree) or is part of the interior of the study region (thus is a focal tree whose growth is of modeled interest).}\label{fig:scbi-spatial-information}
\end{figure}

@@

\caption{Step 3 - Identify all focal and corresponding competitor trees. The dashed circle extends 7.5m away from the focal tree 4 while all 20 competitor trees are within this circle.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}

@@

\caption{Step 4 - Fit model. Posterior distributions of all parameters. For compactness we include only three species.}\label{fig:scbi-posterior-viz}
\end{figure}

@@

This process is repeated for each of the four folds acting as the test
set, then the four RMSE's are averaged to provide a single estimate of
model error. This algorithm is implemented in \texttt{run\_cv()}, which
acts as a wrapper function to both \texttt{comp\_bayes\_lm()} that fits
the model and \texttt{predict()} that returns predicted values.
Furthermore, \texttt{comp\_dist} can now be specified as a vector of
competitive distances to consider.

@@

\caption{Comparison of root mean squared error of models for standard, permuted, and spatially cross-validated error estimates. The dotted lines show observed RMSE while the histograms show the null distribution of RMSE for 49 permutations under the null hypothesis of no competitor species identity effects. The colors indicate whether spatial cross-validation was used or not.}\label{fig:scbi-simulation}
\end{figure}

\hypertarget{conclusion-and-future-work}{%
\section{Conclusion and future work}\label{conclusion-and-future-work}}

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. The package follows the
\texttt{tidy} data design principles, leverages the \texttt{sf} package
for spatial data, and S3 open-oriented model implementation structure.
We hope that the package will increase the use of neighborhood
competition models to better understand what structures plant
competition.

While the package is designed with ForestGEO plot data in mind, we
envision that it can be modified to work on any single large, mapped
forest plot in which at least two measurements of each individual have
been taken. Furthermore, we also envision the package being used more
generally for inventory plot-structure with many spatially separated
plots, e.g.~the US Forest Service Forest Inventory and Analysis plots
\citep{smith_forest_2002}.

We hope to extend the \texttt{forestecology} package's functionality to
account for a larger variety of models for tree growth and corresponding
tests. Some immediate

\begin{quote}
To address the reviewer's critique that the model is not flexible enough
and does not include applications of existing competition theory, we
could add another option for run\_shuffle. In this case the shuffle
would not shuffle identity of competitors which share with the focal
species. The idea being there is a special kind of competition with
conspecifics (because they share your exact resource needs). The result
is two lambda values, one for competition between conspecifics and one
for competition between heterospecifics.
\end{quote}

A recently emerging body of literature suggests that traits of species
rather than species identity themselves better inform competition models
\citep[\citet{lasky_trait-mediated_2014}]{kunstler_competitive_2012}. To
account for such clustering, instead of using Equation \ref{eq:model},
we could use a model of which is a modification of
\citet{kunstler_plant_2016}. Let \(Y_{ik}\) be the growth of the
\(k^{\text{th}}\) individual of species \(i\) and let \(j\) be the
species of competitors in the neighborhood. How much growth changes per
unit of \(\text{BA}\) of all neighbors is quantified by two terms: an
overall competition coefficient \(C\) and an \(i\)-\(j\) species pair
specific coefficient \(c_{ij}\). which depends on the traits of the
\(i\) and \(j\) species. Say we have \(l\) traits and \(t_{li}\) is the
value of trait \(l\) for species \(i\). \(a_{\text{tol},l}\) quantifies
how trait \(l\) affects the way a species tolerates (or is sensitive to)
competition. \(a_{\text{eff},l}\) quantifies how trait \(l\) is how much
trait \(l\) affects how strong a competitor is, how much it effects the
reduction in growth of its neighbors. Finally \(a_{\text{sim},l}\) is
about whether competition is stronger (or weaker) if the two species are
similar. This is based on an idea that two species that are similar use
the same resources and should compete more.

\label{eq:model2} \begin{eqnarray}
y_{ik} &=& \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{j} (C + c_{ij}) \cdot \text{BA}_{j} + \epsilon_{ij}\\
\text{where } c_{ij} &=& \sum_{l} a_{\text{tol},l}\cdot t_{li} + a_{\text{eff},l}\cdot t_{lj} + a_{\text{sim},l}|t_{li} - t_{lj}|\nonumber
\end{eqnarray}

Lastly, the \texttt{forestecology} current uses the \texttt{blockCV}
package behind the scenes to create the spatial blocks acting as folds
for our spatial cross-validation algorithm detailed in Sections
\ref{spatial-information} and \ref{spatial-cross-validation}
\citep{valavi_blockcv_2019}. This back-end functionality could be
substituted with the \texttt{spatialsample} package for spatial
resampling infrastructure; a \texttt{tidymodels} package under active
development as of 2021
\citep[\citet{tidymodels_package}]{spatialsample_package}.
